---
title: "Wine Quality - Data Visualization"
author: "Paria Ighanian"
output:
  
  slidy_presentation: 
    fig.align: "center"
  beamer_presentation: default
date: "2024-01-23"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```



## <span style="color: #336699;">Dataset</span> 

The Wine Quality dataset consists of the chemical properties of both red and white variants of Portuguese ‘Vinho Verde’ wine. It covers features such as alcohol content and acidity levels, alongside quality ratings. This dataset is commonly employed to investigate the relationships between these chemical properties and the perceived quality of the wines. For more details or to access the dataset, you can visit the UCI Machine Learning Repository at [Wine Quality dataset](https://archive.ics.uci.edu/dataset/186/wine+quality).
 

```{r echo=FALSE}
## Libraries

library(ggplot2)
library(dplyr)
library(corrplot)
library(plotly)

```

```{r echo=FALSE}
## Exploring Data

white_wine <- read.csv('winequality_white.csv', sep = ';')
red_wine <- read.csv('winequality_red.csv', sep = ';')

```


```{r echo=FALSE}
red_wine$type <- "Red"
white_wine$type <- "White"

wines <- rbind(red_wine, white_wine)
```


```{r echo=FALSE}
str(wines)
```
## <span style="color: #336699;">Wine Types</span> 

<!--The **Type** feature is a categorical variable that denotes the wine variety, with two possible values: 'Red' and 'White'. Within the dataset, there are 4898 entries classified as 'White' and 1599 entries classified as 'Red'. -->


```{r echo=FALSE}

df_percentage <- wines |>
  group_by(type) |>
  summarise( count = n(),percentage = n() / nrow(wines) * 100) |>
  mutate(percentage = round(percentage, 1))


plot_01 <- ggplot(df_percentage, aes(x = type, y=percentage, fill = type, text= paste0("Count: ", count))) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)),
            position = position_stack(vjust = 0.5))+
  labs(x = NULL,
       y = NULL)+
  scale_y_continuous(labels = scales::percent_format(scale = 1))+
  theme(axis.title.y = element_text(margin = margin(r = 10)),
        axis.text.x = element_text(size = 12),  
        legend.position = "none",
        plot.title = element_text(hjust = 0.5, margin = margin(b = 20)))+

  scale_fill_manual(values = c("Red" = "#B94155", "White" = "#E4E06D")) +
  ggtitle("Wine Type Distribution")

plot_01_ly <- ggplotly(plot_01, tooltip = c("text")) |> config(displayModeBar = FALSE)
plot_01_ly
```


## <span style="color: #336699;">Quality Distribution</span> 

<!--The **quality** is a discrete rating that represents a wine quality score within the range of 3 to 9.-->


```{r echo=FALSE}
quality_distribution_by_type <- wines |>
  group_by(type, quality) |>
  summarise(count = n()) |>
  mutate(percentage = round(count / sum(count) * 100, 1)) 


plot_02 <-ggplot( quality_distribution_by_type, aes(x = as.factor(quality), y = percentage, fill = type, text= paste0(percentage, "%"))) +
  geom_bar(stat = "identity", alpha = 0.7) +
  
  labs(x = NULL,
       y = NULL, title = "Distribution of Wine Quality") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  
  scale_fill_manual(values = c("Red" = "#B94155", "White" = "#E4E06D")) +
  theme(axis.title.y = element_text(margin = margin(r = 10)),
        axis.text.x = element_text(size = 12),  
        legend.position = "none",
        plot.title = element_text(hjust = 0.5),
        plot.margin = margin(t = 20)) +
  

  facet_wrap(~type, ncol=2)


plot_02_ly <- ggplotly(plot_02, tooltip = c("text")) |> config(displayModeBar = FALSE)

plot_02_ly


```



## <span style="color: #336699;">Quality vs Alcohol</span> 



```{r echo=FALSE}

plot_white <- ggplot(aes(factor(quality), alcohol), data = white_wine) +
  geom_boxplot(fill = "#E4E06D") +
  geom_jitter(alpha = 0.05) 

plot_red <- ggplot(aes(factor(quality), alcohol), data = red_wine) +
  geom_boxplot(fill = "#B94155") +
  geom_jitter(alpha = 0.05) 

plot_white_ly <- ggplotly(plot_white,  tooltip = 'none') |> config(displayModeBar = FALSE)
plot_red_ly <- ggplotly(plot_red,  tooltip = 'none') |> config(displayModeBar = FALSE)


combined_plot <- subplot(plot_white_ly, plot_red_ly, nrows = 1)



combined_plot <- layout(
  combined_plot,
  yaxis = list(title = 'Alcohol (%)'),
  title = list(
    text = "Correlation of Alcohol and Quality",
    margin = margin(b = 20)  
  ),
  margin = list(t = 50)  
)

combined_plot


```

## <span style="color: #336699;">Alcohol and Density's Role in Wine Quality</span> 


```{r echo=FALSE}
plot_04 <- ggplot(data = wines, aes(x = alcohol, y = density, color = factor(quality))) +
  geom_point(alpha = 0.6, position = position_jitter(h = 0), size = 3, stroke = 0.2) +
  geom_smooth(method = 'lm', se = FALSE, color = "black", linetype = "dashed", size = 1) +  
  coord_cartesian(ylim = c(0.987, 1.003)) +
  scale_colour_brewer(palette = "Set1") +
  labs(x = "Alcohol(%)", y = "Density", color = "Quality") +
  theme(axis.title.y = element_text(margin = margin(r = 10)),
        axis.text.x = element_text(size = 12),  
        plot.title = element_text(hjust = 0.5),
        plot.margin = margin(t = 20)
        ) +
  
  
  facet_wrap(~ type) 
plot_04_ly <-ggplotly(plot_04,  tooltip = 'none') |> config(displayModeBar = FALSE)
plot_04_ly <- layout(plot_04_ly,
    title = list(
    text = "Density and Alcohol Effects on Quality Grading",
    margin = margin(b = 20)  
  ),
    margin = list(l = 70)  
  )

plot_04_ly
```

## <span style="color: #336699;">Residual Sugar and Density's Role in Wine Quality</span> 

```{r echo=FALSE}

plot_05 <- ggplot(data = wines, aes(x = residual.sugar , y = density, color = factor(quality))) +
  geom_point(alpha = 0.6, position = position_jitter(h = 0), size = 3, stroke = 0.2) +
  geom_smooth(method = 'lm', se = FALSE, color = "black", linetype = "dashed", size = 1) +  
  coord_cartesian(ylim = c(0.987, 1.003)) +
  scale_colour_brewer(palette = "Set1") +
  labs(x = "Residual Sugar ", y = "Density", color = "Quality") +
  theme(axis.title.y = element_text(margin = margin(r = 10)),
        axis.text.x = element_text(size = 12),  
        plot.title = element_text(hjust = 0.5),
        plot.margin = margin(t = 20)
        ) +
  
  
  facet_wrap(~ type) 
plot_05_ly <-ggplotly(plot_05,  tooltip = 'none') |> config(displayModeBar = FALSE)
plot_05_ly <- layout(plot_05_ly,
    title = list(
    text = "Density and Residual Sugar Effects on Quality Grading",
    margin = margin(b = 30)  
  ),
    margin = list(l = 70)  
  )

plot_05_ly


```



